***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**

### 2015-10-15 Schloss_Cluster_PeerJ_2015.Rmd
Edit Abstract and Introduction

**Background.** 16S rRNA gene sequences are routinely assigned to operational taxonomic units (OTUs) that are then used to analyze complex microbial communities. A number of methods have been employed to carry out the assignment of 16S rRNA gene sequences to OTUs leading to confusion over which method is the best. A recent study has suggested that a clustering method should be selected based on its ability to generate stable OTU assignments as additional sequences are added to the dataset rather than the ability of the method to properly represent the distances between the sequences. It was thus necessary to replicate the previous analysis and assess the quality of the OTU assignments.

**Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering and two reference-based methods including the open and closed-reference methods. By varying the number of sequences sampled from each of two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the stability of OTU assignments and their quality.

**Results.** Our analysis demonstrated that the stability of OTU assignments does not predict to the ability of the methods to reflect the distance between the sequences. We found that the average linkage and the distance and abundance-based greedy clustering methods out performed the open and closed-reference methods. Depending on the dataset being analyzed, the average linkage or greedy clustering algorithms provided the best dataset. Further exploration of the reference-based methods indicated that when using USEARCH to identify the closest reference the results were sensitive to the order of the reference sequences. When USEARCH was replaced by VSEARCH the quality and stability of the OTU assignments by the reference-based methods were improved. Finally, we demonstrated that for the greedy algorithms VSEARCH produced assignments that were comparable to those produced by USEARCH making VSEARCH a viable free and open source alternative to USEARCH.

@@

The ability to affordably generate millions of 16S rRNA gene sequences has allowed microbial ecologists to thoroughly characterize the microbial community composition of hundreds of samples. To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins. These bins, commonly known as operational taxonomic units, are  used to compare the biodiversity contained within and between different samples [@schloss2011assessing]. Such comparisons have enabled researchers to characterize the microbiota associated with the human body [@human2012structure], soil [@shade2013streptomycin], aquatic ecosystems [@gilbert2012defining], and numerous other environments. Within the field of microbial ecology a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%. One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project. However, with the dissemination of clustering methods within software such as mothur [@schloss2009introducing], QIIME [@Caporaso2010], and other tools [@edgar2010search, @sun2009esprit, @10.7717/peerj.593], it is important to understand how different clustering methods implement this conventional OTU threshold. Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs. Broadly speaking, three approaches have been developed to assign sequences to OTUs.

The first approach has been referred to as phylotyping [@schloss2011assessing] or closed reference clustering [@navas2013advancing]. This approach involves comparing sequences to a curated database and then clustering sequences together that are similar to the same reference sequence. Reference-based clustering methods suffer when the reference sequences do not reflect the composition of the community. If a large fraction of sequences are novel, then they cannot be assigned to an OTU. In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene. Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences. Because of this, defining OTUs in the closed-reference approach is complicated because two sequences might be 97% similar to the same reference sequence; however, they may only be 94% similar to each other. The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered. A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@schloss2011assessing]. Ultimately, this approach is an example of supervised classification.

The second approach has been referred to as distance-based [@schloss2011assessing] or *de novo* clustering [@navas2013advancing]. In this method, the distance between sequences is used to cluster sequence rather than the distance to a reference database. In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences. The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences. If error rates can be reduced through stringent quality control measures then these problems can be overcome [@Kozich2013]. As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@edgar2010search, @sun2009esprit, @10.7717/peerj.593]. Previous comparisons of hierarchical and heuristic methods have shown that the average neighbor method, a hierarchical method, outperformed the other *de novo* clustering methods. An early analysis using 16S rRNA gene sequence data generated using the 454 sequencing platform found that if these methods were applied to the same dataset using a subset of the sequences then a different number of OTUs were observed compared to clustering the full dataset and rarefying to the same number of sequences. Because the initial subsampling of the dataset was not repeated in the earlier analysis, it is unclear whether the result was product of the method or sampling. One explanation could be that the clusters are getting better with additional sampling. Another observed problem with *de novo* clustering is that because it is necessary to break ties when assigning sequences to OTUs, clustering the same data multiple times may result in different clustering assignments. Whether the differences in assignments is meaningful is unclear; however the variation in results could represent equally valid clustering of the data. The strength of *de novo* clustering is its independence of references for carrying out the clustering step. After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@schloss2011assessing]. For this reason, *de novo* clustering has been preferred across the field.

The third approach is a hybrid of the closed-reference and *de novo* approaches. This approach has been called open-reference clustering. This method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference [@navas2013advancing]. This method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions generated by both approaches poses a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels [@schloss2011assessing]. For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average neighbor method. Those sequences that did not classify to a known family would also be clustered using the average neighbor method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group (e.g. each family) is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.

The growth in options for assigning sequences using each of these three broad approaches has been considerable. It has been difficult to objectively assess the quality of OTU assignments. Some have focused on the time and memory required to process a dataset (REFS). These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering. Others have attempted to judge the quality of an method by its ability to generate data that parallels classification data. This approach is problematic because bacterial taxonomy often reflects the biases amongst bacterial systematicists and the rates of evolution across lineages are not the same. We recently proposed an approach for evaluating OTU assignments using the distance between sequences (REF). Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives. Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives. Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and ratios using the Matthew's correlation coefficient (MCC; REF). That analysis focused on assessing *de novo* clustering methods and found that the average neighbor method outperformed the other hierarchical and heuristic methods.

A recent analysis by He and colleagues (REF) attempted to characterize the three general clustering approaches by focusing on what they called stability. They defined stability as the ability of an method to provide the same clustering on a subset of the data as was found in the full dataset. Related to this, the authors expressed concern that because some methods use a random number generator to break ties there may be further instability between executions of method. Their concept of stability does not account for the accuracy of the clustering and instead focuses on the precision of the clustering. A method may be very precise, but low in accuracy. In the current analysis, we attempted to assess the accuracy and precision of the various clustering methods. Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of accuracy. In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.

>>>>

**Background.** 16S rRNA gene sequences are routinely assigned to operational taxonomic units (OTUs) that are then used to analyze complex microbial communities. A number of methods have been employed to carry out the assignment of 16S rRNA gene sequences to OTUs leading to confusion over which method is the best. A recent study suggested that a clustering method should be selected based on its ability to generate stable OTU assignments that do not change as additional sequences are added to the dataset. In contrast, we contend that the ability of the method to properly represent the distances between the sequences is more important.

**Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering and two reference-based methods including the open and closed-reference methods. By varying the number of sequences sampled from each of two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the quality and stability of OTU assignments.

**Results.** The stability of OTU assignments did not reflect the quality of the assignments. Depending on the dataset being analyzed, the average linkage and the distance and abundance-based greedy clustering methods generated more robust OTUs than the open and closed-reference methods. Further interrogation of the reference-based methods indicated that when USEARCH is used to identify the closest reference, the OTU assignments were sensitive to the order of the reference sequences. When USEARCH was replaced by VSEARCH the quality and stability of the OTU assignments by the reference-based methods were improved. Finally, we demonstrated that for the greedy algorithms VSEARCH produced assignments that were comparable to those produced by USEARCH making VSEARCH a viable free and open source alternative to USEARCH.

@@

The ability to affordably generate millions of 16S rRNA gene sequences has allowed microbial ecologists to thoroughly characterize the microbial community composition of hundreds of samples. To simplify the complexity of these large datasets, it is helpful to cluster sequences into meaningful bins. These bins, commonly known as operational taxonomic units (OTUs), are  used to compare the biodiversity contained within and between different samples [@schloss2011assessing]. Such comparisons have enabled researchers to characterize the microbiota associated with the human body [e.g. @human2012structure], soil [e.g. @shade2013streptomycin], aquatic ecosystems [e.g. @gilbert2012defining], and numerous other environments. Within the field of microbial ecology, a convention has emerged where sequences are clustered into OTUs using a threshold of 97% similarity or a distance of 3%. One advantage of the OTU-based approach is that the definition of the bins is operational and can be changed to suit the needs of the particular project. However, with the dissemination of clustering methods within software such as mothur [@schloss2009introducing], QIIME [@Caporaso2010], and other tools [@edgar2010search, @sun2009esprit, @10.7717/peerj.593], it is important to understand how different clustering methods implement this conventional OTU threshold. Furthermore, it is necessary to understand how method choice affects the precision and accuracy of assigning sequences to OTUs. Broadly speaking, three approaches have been developed to assign sequences to OTUs.

The first approach has been referred to as phylotyping [@schloss2011assessing] or closed reference clustering [@navas2013advancing]. This approach involves comparing sequences to a curated database and then clustering sequences into the same OTU that are similar to the same reference sequence. Reference-based clustering methods suffer when the reference does not adequately reflect the biodiversity of the community. If a large fraction of sequences are novel, then they cannot be assigned to an OTU. In addition, the reference sequences used in this application are selected because they are less than 97% similar to each other over the full length of the gene; however, it is known that the commonly used variable regions within the 16S rRNA gene do not evolve at the same rate as the full-length gene. Thus, a sequence representing a fragment of the gene may be more than 97% similar to multiple reference sequences. Therefore, defining OTUs in the closed-reference approach is complicated because although two sequences might be 97% similar to the same reference sequence, they may only be 94% similar to each other. A subtle alternative to this approach is to use a classifier to assign a taxonomy to each sequence so that sequences can be clustered at a desired level within the Linnean taxonomic hierarchy [@schloss2011assessing]. The strength of the reference based approach is that the methods are generally fast, scaling linearly with the number of sequences being clustered.

The second approach has been referred to as distance-based [@schloss2011assessing] or *de novo* clustering [@navas2013advancing]. In this method, the distance between sequences is used to cluster sequences into OTUs rather than the distance to a reference database. In contrast to the efficiency of closed-reference clustering, the speed of hierarchical *de novo* clustering methods scale quadratically with the number of unique sequences. The expansion in sequencing throughput combined with sequencing errors inflates the number of unique sequences resulting in the need for large amounts of memory and time to cluster the sequences. If error rates can be reduced through stringent quality control measures then these problems can be overcome [@Kozich2013]. As an alternative, heuristics have been developed to approximate the clustering of hierarchical methods [@edgar2010search, @sun2009esprit, @10.7717/peerj.593]. One critique of *de novo* approaches is that OTU assignments are sensitive to the input order of the sequences [REFS]. Whether the differences in assignments is meaningful is unclear; however, the variation in results could represent equally valid clustering of the data. The strength of *de novo* clustering is its independence of references for carrying out the clustering step. After clustering, the classification of each sequence can be used to obtain a consensus classification for the OTU [@Schloss2011]. For this reason, *de novo* clustering has been preferred across the field.

The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@Navas2013]. Open-reference clustering method involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference. In theory, this method should exploit the strengths of both closed-referene and *de novo* clustering; however, the different OTU definitions employed by both approaches poses a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assigned to OTUs within those levels using the average linkage method [@Schloss2011]. For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold. Those sequences that did not classify to a known family would also be clustered using the average linkage method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches.

The growth in options for assigning sequences using each of these three broad approaches has been considerable. It has been difficult to objectively assess the quality of OTU assignments. Some have focused on the time and memory required to process a dataset (REFS). These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the clustering. Others have attempted to judge the quality of an method by its ability to generate data that parallels classification data (REFS). This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists. Furthermore, it is well known that the rates of evolution across lineages are not the same (REFS). We recently proposed an approach for evaluating OTU assignments using the distance between sequences (REF). Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives. Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives. Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the Matthew's correlation coefficient (MCC; REF). This is an objective approach to assessing the quality of the OTU assignments.

A recent analysis by He and colleagues (REF) attempted to characterize the three general clustering approaches by focusing on what they called stability. They defined stability as the ability of an method to provide the same clustering on a subset of the data as was found in the full dataset. Their concept of stability did not account for the accuracy of the OTU assignments and instead focused on the precision of the assignments. A method may be very precise, but low in accuracy. In the current analysis, we assessed the accuracy and precision of the various clustering methods. Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of accuracy. In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.
